---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TabPFN

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/TabPFN)](https://CRAN.R-project.org/package=TabPFN)
[![R-CMD-check](https://github.com/topepo/TabPFN/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/topepo/TabPFN/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/topepo/TabPFN/branch/main/graph/badge.svg)](https://app.codecov.io/gh/topepo/TabPFN?branch=main)
<!-- badges: end -->

TabPFN, meaning prior fitted networks for tabular data, is a deep-learning model. See:

- [_Transformers Can Do Bayesian Inference_](https://arxiv.org/abs/2112.10510) (arXiv, 2021)
- [_TabPFN: A Transformer That Solves Small Tabular Classification Problems in a Second_](https://arxiv.org/abs/2207.01848) (arXiv, 2022)
- [_Accurate predictions on small data with a tabular foundation model_](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C7&q=%22Accurate+predictions+on+small+data+with+a+tabular+foundation+model%22) (Nature, 2025)

This R package is a wrapper of the [Python library](https://github.com/PriorLabs/TabPFN) via reticulate. It has an idiomatic R syntax using standard S3 methods. 

## Installation

You can install the development version of TabPFN like so:

```{r}
#| eval: false
require(pak)
pak(c("topepo/TabPFN"), ask = FALSE)
```

You'll need a Python virtual environment to access the underlying library. After installing the R package, you could use this code to create an appropriate environment: 

```{r}
#| eval: false
library(reticulate)

python <-install_python()

virtualenv_install(
	"r-tabpfn",
	packages = c("numpy", "tabpfn"),
	python = python
)

# check the install:
py_config()
py_list_packages(envname = "r-tabpfn", type = "virtualenv")
```


## Example


```{r}

```

On starting the TabPFN, it will see if the python packages are installed. 

```{r}
#| label: tab-start-up

# Perhaps start by setting your virt env: 
# reticulate::use_virtualenv("~/.virtualenvs/your-envir-name")

library(TabPFN)
```

To fit a model: 

```{r}
#| label: mtcars
reg_mod <- TabPFN(mtcars[1:25, -1], mtcars$mpg[1:25])
reg_mod
```

In addition to the x/y interface shown above, there are also formula and recipes interfaces. 

Prediction follows the usual S3 `predict()` method: 

```{r}
#| label: mtcars-pred
predict(reg_mod, mtcars[26:32, -1])
```

While TabPFN isn’t a tidymodels package, it follows their prediction convention: a data frame is always returned with a standard set of column names. 

For a classification model, the outcome should always be a factor vector. For example, using these data from the modeldata package: 

```{r}
#| label: cls
#| results: none
library(modeldata)
library(ggplot2)

two_cls_train <- parabolic[1:400,  ]
two_cls_val   <- parabolic[401:500,]
grid <- expand.grid(X1 = seq(-5.1, 5.0, length.out = 25), 
                    X2 = seq(-5.5, 4.0, length.out = 25))

cls_mod <- TabPFN(class ~ ., data = two_cls_train)

grid_pred <- predict(cls_mod, grid)
grid_pred
```

The fit looks fairly good when shown with out-of-sample data: 

```{r}
#| label: boundaries
#| fig.width: 5
#| fig.height: 4
#| fig.align: "center"
#| out.width: 50%

cbind(grid, grid_pred) |>
 ggplot(aes(X1, X2)) + 
 geom_point(data = two_cls_val, aes(col = class, pch = class), 
            alpha = 3 / 4, cex = 3) +
 geom_contour(aes(z = .pred_Class1), breaks = 1/ 2, col = "black", linewidth = 1) +
 coord_equal(ratio = 1)
```

## Code of Conduct
  
Please note that the TabPFN project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
