---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TabPFN

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/TabPFN)](https://CRAN.R-project.org/package=TabPFN)
<!-- badges: end -->

The goal of TabPFN is to ...

## Installation

You can install the development version of TabPFN like so:

```{r}
#| eval: false
require(pak)
pak(c("topepo/TabPFN"), ask = FALSE)
```

## Example


The package requires a virtual environment to be created and registered with reticulated. If you don't have one, you can create one. First, load the reticulate package: 

```{r}
require(reticulate)
```

and this code can be used to create an environment and install the relevant packages: 

```{r}
#| label: venv-setup
#| eval: false

virtualenv_create(
	"r-tabpfn",
	packages = c("numpy", "tabpfn"),
	python_version = "<3.12"
)
```

then tell reticulate to use it: 

```{r}
use_virtualenv("~/.virtualenvs/r-tabpfn")
```

On starting the TabPFN, it will see if the python packages are installed. 

```{r}
#| label: tab-start-up
library(TabPFN)
```

There is also a function to check too: 

```{r}
#| lable: check-python

TabPFN:::check_py_packages()
```

To fit a model: 

```{r}
#| label: mtcars
reg_mod <- TabPFN(mtcars[1:25, -1], mtcars$mpg[1:25])
reg_mod
```

In addition to the x/y interface shown above, there are also formula and recipes interfaces. 

Prediction follows the usual S3 `predict()` method: 

```{r}
#| label: mtcars-pred
predict(reg_mod, mtcars[26:32, -1])
```

While TabPFN isn’t a tidymodels package, it follows their prediction convention: a data frame is always returned with a standard set of column names. 

For a classification model, the outcome should always be a factor vector. For example, using these data from the modeldata package: 

```{r}
#| label: cls
require(modeldata)
require(ggplot2)

two_cls_train <- parabolic[1:400,  ]
two_cls_val   <- parabolic[401:500,]
grid <- expand.grid(X1 = seq(-5.1, 5.0, length.out = 25), 
                    X2 = seq(-5.5, 4.0, length.out = 25))

cls_mod <- TabPFN(class ~ ., data = two_cls_train)

grid_pred <- predict(cls_mod, grid)
grid_pred
```

The fit looks fairly good when shown with out-of-sample data: 

```{r}
#| label: boundaries
#| fig.width: 5
#| fig.height: 4
#| fig.align: "center"
#| out.width: 50%

cbind(grid, grid_pred) |>
 ggplot(aes(X1, X2)) + 
 geom_point(data = two_cls_val, aes(col = class, pch = class), 
            alpha = 3 / 4, cex = 3) +
 geom_contour(aes(z = .pred_Class1), breaks = 1/ 2, col = "black", linewidth = 1) +
 coord_equal(ratio = 1)
```

## AutoTabPFN
AutoTabPFN (called "TabPFN (PHE)" in the original [TabPFN paper](https://www.nature.com/articles/s41586-024-08328-6)) is an ensemble version of TabPFN, that automatically runs a hyperparameter search and build an ensemble of TabPFN models. It is slower to train and use, but on average yields better predictions.

Using AutoTabPFN requires that you install the [tabpfn-community.post_hoc_ensembles package](https://github.com/PriorLabs/tabpfn-community/tree/main). At the time of writing, this is only available from GitHub, and can be installed from a terminal using:

```{bash}
cd ~/.virtualenvs/r-tabpfn
source bin/activate
git clone https://github.com/PriorLabs/tabpfn-community
pip install -e tabpfn-community[post_hoc_ensembles]
```

To train an AutoTabPFN ensemble, simply use `AutoTabPFN()` analogously to how `TabPFN()` is used. You can then use `predict()` for predictions.

```{r}
#| label: mtcars-auto
reg_mod <- AutoTabPFN(mpg ~ ., data = mtcars[1:25, ])
predict(reg_mod, mtcars[26:32, -1])
```



## Code of Conduct
  
Please note that the TabPFN project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
